---
title: "PaleoFidelity"
author: "Michal Kowalewski"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PaleoFidelity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The package "PaleoFidelity" carries out fidelity analyses (live-dead comparisons) and is intended for compositional data that represent specimen counts of taxa with separate tallies for live and dead components of each sample. The package can be used for a single site (one live-dead comparison) or sets of sites. A grouping factor can be provided to split sites into two or more subsets and carry out comparative analyses across groups (habitats, regions, etc.). An additional grouping factor that splits variables/columns (e.g., species) into subsets can be used to assess how fidelity varies across variable groups (e.g., phyla, functional categeories, or taphonomic groups).

Currently, the following analyses (discussed in more detail below) can be performed:

1. __FidelitySummary__ function to pre-assess data prior to any other fidelity analysis.
2. __FidelityEst__ function performs compositional fidelity analysis based on correlation/similarity measures adjusted for sampling coverage.
3. __FidelityDiv__ function performs alpha diversity and evenness analysis (live-dead offsets in standardized diversity and evenness).
4. __FidelityBeta__ function assesses live-dead fidelity across sites (offsets in compositional heterogeneity or gradients across sites).
5. __SJPlot__ plots a fidelity cross-plot of correlation and similarity measures of fidelity 
6. __AlphaPlot__ plots a fidelity cross-plot of live-dead offsets in alpha diversity and eveness
6. __LDPlot__ plots a comparative live-dead plot for the most common taxa

## Installing PaleoFidelity Package

Currently, the package is available at github. Use the following statements to install and upload it:
**install.packages('devtools')**
**library(devtools)**
**devtools::install_github('mjkowalewski/PaleoFidelity', build_vignettes = TRUE)**
**library(PaleoFidelity)**


## Example dataset

An example dataset is included in the package to illustrate live-dead data suitable for fidelity analyses.

```{r data example}
library(PaleoFidelity)
str(FidData) # check the structure of the example dataset
```

The example dataset is assembled as a single list with 4 items, but data can be separate objects. Live and dead datasets must be formatted as matrices (use __as.matrix__ function to convert) and their dimensions must match exactely. The two categorical variables that can be included need to be univariate factors.

For a single live-dead analysis, format 'live' and 'dead specimen counts as one-row matrices (e.g., rbind(live), rbind(dead)). The same applies when extracting one sample pair from a larger matrix.
For example:  rbind(live[7,]) and rbind(dead[7,]).  

## FidelitySummary Function

A function _FidelitySummary_ assesses data for format compliance and numerical adequacy for carrying out a meaningful fidelity analysis. The function also allows to filter rare taxa and small samples. Note that this function is called by other PaleoFidelity functions and the data filters should be set in those functions before the analysis is carried out. Whereas _FidelitySummary_ is a convenient way to pre-assess data and determine filters, it does not produce any analytical outcomes.

When the function is applied to the example dataset in its default form (but with report=TRUE), a report, notes, and one warning are printed.

```{r fidelity summary function}
FidelitySummary(live=FidData$live, dead=FidData$dead, gp=FidData$habitat, report=TRUE)
```

Note that the report indicates that there are 44 samples and 202 taxa with at least one observation and nothing was removed from the data (otherwise, some columns and/or rows would have been be removed and the number of rows and columns removed would have been reported). You can also learn now that the grouping factor included 2 levels and both levels contained more than one sample making those levels 'useful' for statistical analyses. However, a warning indicates that some samples are small (n < 30) and smallest sample sizes are listed in the report as 17 for live and 20 for dead, respectively.

The function can be used to explore the consequence of applying the filters. For example, let's remove all samples with n < 30. 

```{r fidelity summary function part 2}
FidelitySummary(live=FidData$live, dead=FidData$dead, gp=FidData$habitat,
                report=TRUE, n.filters=30)
```

Note that no taxa were lost (i.e., the sites that were removed did not contain any unique species), 6 sites were removed (the number of sites is now 38 compared to 44 originally present), and the grouping factor still includes more than 1 site per level. This, the n.filter=30 did not reduce data dimensionality at all and all factor groups (levels) still retained at least two observations/sites. How about n.filters=100?

```{r fidelity summary function part 3}
FidelitySummary(live=FidData$live, dead=FidData$dead, gp=FidData$habitat,
                report=TRUE, n.filters=100)
```

Now, 12 taxa were lost and only 21 sites were retained. Perhaps n.filters=100 is an unreasonably high n.filter?

Once the appropriate filters are determined, they can be implemented in other fidelity functions. 

## FidelityEst Function

A function _FidelityEst_ estimates compositional fidelity measures by comparing two matching matrices (live and dead) with community abundance data. The function returns  fidelity estimates for indiviudal sites and for groups of sites if 'gp' factor is provided. The function uses a bivariate approach and produces two estimates of compositional fidelity:
1. $x - a measure of correlation/association: spearman (default), kendall, or pearson.
2. $y - an abundance-based indices of similarity such as bray (default) or jaccard-chao.

Because many of those fidelity measures tend to be sensitive to unbalanced sampling, _FidelityEst_ function attempts to correct sampling bias by assessing data-specific biases in correlation/similarity measures. The bias is estimated using a resampling protocol under the perfect fidelity (PF) model, in which pooled (live + dead) counts are randomly partitioned into replicate pairs of samples (using sample sizes of original samples), thus creating sample pairs derived from a single underlying rank abundance distribution of species (i.e., perfect fidelity). For an unbiased estimator, the resampled fidelity measures should indicate perfect fidelity (e.g., Spearman rho = 1). The offset between the expected value of 1 and the observed value (1 - PF) provides a data-specific estimate of sampling bias. The adjusted fidelity measure for a given live-dead coparison is then given by Adjusted = Observed + (1 - PF). Replicate resampling produces a distribution of PF values and resulting adjusted fidelity measures, from which confidence intervals and signifiance tests can be derived.

When the data include multiple sites that can be grouped into subsets, _FidelityEst_ will report mean fidelity estimates for total data and each group. Two measures are reported: a correlation measure (default=Spearman) and a similarity measure (default=Bray). Note also that correlation can be computed with double zeroes included or excluded. In the example below, double zeroes were kept (rm.zero=F), which is the recommended approach.

The function returns a list and uncorrected measures of correlation and similarity for each live-dead sample pair (stored as $x and $y, respectively).

```{r fidelity estimates}
out1 <- FidelityEst(live=FidData$live, dead=FidData$dead, gp=FidData$habitat, n.filters=30, rm.zero=F, iter=999)
out1$x
out1$y
```

The function returns also the adjusted fidelity measures ($xc, $yc) corrected for sampling coverage bias. The bias correction is based on a resampling model simulating 'Perfect Fidelity' in which departures from perfect fidelity are due solely to sampling a perfectly congruent live-dead system. In addition, mean values for all samples and sample groups are also reported (if factor 'gp' is provided).

```{r fidelity estimates outputs}
out1$xc # adjusted correlation measure summary
out1$yc # adjusted similarity measure summary
```



```{r fidelity estimates part 2}
out1$x.stats
out1$y.stats
```

To explore data visually use _SJPlot_ function. In its default form, this function produces a bubble plot with points scaled by the sample size of the smaller of the two (live or dead) compared samples. For the default setting only adjusted values are plotted for each sample pair.

## SJPlot Function

```{r classic fidelity plot, fig.width=7.5, fig.height=4}
par(mar=c(4, 4, 0.5, 0.5))
SJPlot(out1, gpcol=c('aquamarine3', 'coral3'), legend.cex=0.8)
```

Set bubble=FALSE to plot unscaled symbols. Set unadjF=T to plot uncorrected estimates (adjF=F suppresses plotting adjusted measures and CI=F suppresses confidence intervals).

```{r classic fidelity plot 2, fig.width=7.5, fig.height=4}
par(mar=c(4, 4, 0.5, 0.5))
SJPlot(out1, gpcol=c('aquamarine3', 'coral3'), bubble=F, unadj=T, legend.cex=0.8)
```

## FidelityDiv Function

To evaluate live-dead fidelity in terms of alpha diversity and evenness, use _FidelityDiv_ function. To visualize those results call function AlphaPlot. When sites are categorized into groups, diversity analyses include by-group analyses and by-group plotting is returned by AlphaPlot.

The function will report the offsets in alpha diversity and evenness (stored as $x and $y, respectively) for each live-dead comparisons. For example,

```{r alpha diversity}
out3 <- FidelityDiv(FidData$live, FidData$dead, iter=1000)
out3$x
out3$y
```

The returned object also mean offsets across all samples and groups of samples (if the factor 'gp' is provided). Significance values are also reported for the null hypothesis: offset = 0.

```{r alpha diversity 2}
out4 <- FidelityDiv(FidData$live, FidData$dead, FidData$habitat, iter=1000)
out4$xmean
out4$ymean
out4$xgp
out4$ygp
out4$p.values
out4$p.gps
```

# AlphaPlot Function

```{r plot alpha 2, fig.width=7.5, fig.height=4}
out3 <- FidelityDiv(FidData$live, FidData$dead, FidData$habitat, CI=0.95, iter=1000)
par(mar=c(4, 4.5, 0.5, 0.5))
AlphaPlot(out3, col.gp=c('aquamarine3', 'coral3'), bgpt='beige', pch=22, legend.cex=0.8)
```
